---
title: "REPEL nowcast eval"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 13, fig.height = 10)
devtools::load_all()
library(ggplot2)
library(kableExtra)
library(DALEX)
library(knitr)
library(forcats)
library(yardstick)
library(cowplot)
library(DALEX)
library(rsample)

oie_high_importance_diseases <- get_oie_high_importance_diseases()

model_object <-  nowcast_boost_model(
  disease_status_model = aws.s3::s3readRDS(bucket = "repeldb/models", object = "boost_mod_disease_status.rds"),
  cases_model = aws.s3::s3readRDS(bucket = "repeldb/models", object = "boost_mod_cases.rds"))

cbp1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# Load model
boost_mod_disease_status <- model_object$disease_status_model
boost_mod_disease_status_xg <- pull_workflow_fit(boost_mod_disease_status)

# Get training data
boost_dat_disease_status <- pull_workflow_mold(boost_mod_disease_status)$predictors %>% 
  as.matrix()
boost_dat_out_disease_status <- pull_workflow_mold(boost_mod_disease_status)$outcomes %>% 
  mutate(disease_status = as.integer(disease_status == 1)) %>% 
  pull(disease_status)

# Load model
boost_mod_cases <-  model_object$cases_model
boost_mod_cases_xg <- pull_workflow_fit(boost_mod_cases)

# Get training data
boost_dat_cases <- pull_workflow_mold(boost_mod_cases)$predictors %>%
  as.matrix()
boost_dat_out_cases <- pull_workflow_mold(boost_mod_cases)$outcomes %>%
  pull(cases)
```

## Variable importance
<details>
<summary>disease status overall variable importance</summary>
```{r ds-var-importance}

# Variable importance
importance_disease_status <- vip::vip(boost_mod_disease_status_xg, num_features = 12)
importance_disease_status$data <- 
  importance_disease_status$data %>% 
  mutate(Variable = str_replace(Variable, "lag1", "(6 months)")) %>% 
  mutate(Variable = str_replace(Variable, "lag2", "(12 months)")) %>% 
  mutate(Variable = str_replace(Variable, "lag3", "(18 months)")) %>% 
  mutate(Variable = str_replace_all(Variable, "_", " ")) %>% 
  mutate(Variable = recode(Variable, 
                           "ever in country given taxa" = "disease ever in country (within given taxa)",
                           "ever in country any taxa" = "disease ever in country (within any taxa)",
                           "cases lag sum border countries" = "cases in border countries (prev. 18 months)",
                           "cases (6 months) missing" = "missing value: cases (6 months)",
                           "cases (18 months) missing" = "missing value: cases (18 months)",
                           "log gdp per capita" = "GDP per capita"
                           
  ))

importance_disease_status + 
  ggthemes::theme_fivethirtyeight(base_size = 22) +
  theme(axis.title = element_text(size = 14)) 

```
</details>


<details>
<summary>cases overall variable importance</summary>
```{r cs-var-importance}

# Variable importance
importance_cases <- vip::vip(boost_mod_cases_xg, num_features = 12)
importance_cases$data <- 
  importance_cases$data %>% 
  mutate(Variable = str_replace(Variable, "lag1", "(6 months)")) %>% 
  mutate(Variable = str_replace(Variable, "lag2", "(12 months)")) %>% 
  mutate(Variable = str_replace(Variable, "lag3", "(18 months)")) %>% 
  mutate(Variable = str_replace(Variable, "log_", "")) %>% 
  mutate(Variable = str_replace_all(Variable, "_", " ")) %>% 
  mutate(Variable = recode(Variable, 
                           "cases lag sum border countries" = "cases in border countries (prev. 18 months)",
                           "gdp per capita" = "GDP per capita",
                           "veterinarians per taxa" = "veterinarians per taxa",
                           "taxa birds" = "taxa: birds",
                           "disease mycoplasma infection" = "disease: mycoplasma infection",
                           "country iso3c IRN" = "country: Iran"
  ))

importance_cases + 
  ggthemes::theme_fivethirtyeight(base_size = 22) +
  theme(axis.title = element_text(size = 14)) 

```
</details>

<details>
<summary>disease status partial dependency of lag vars by disease</summary>
```{r ds-lag-pd-by-disease}

lag_vars <- c("disease_status_lag1", "disease_status_lag2", "disease_status_lag3")

ds_pd_by_disease <-  map_dfr(oie_high_importance_diseases, function(disease){
  
  col <- paste("disease", disease, sep = "_")
  if(col %in% colnames(boost_dat_disease_status)){
    which_disease <- which(boost_dat_disease_status[,col] == 1)
    
    if(length(which_disease)){
      explainer_disease_status_sub <- DALEX::explain(
        model = boost_mod_disease_status_xg,
        data = boost_dat_disease_status[which_disease,],
        y = boost_dat_out_disease_status[which_disease],
        predict_function = predict_raw,
        label = "disease status",
        verbose = FALSE
      )
      
      map_dfr(lag_vars, function(var){
        var_resp <- model_profile(explainer_disease_status_sub, 
                                  variable =  var , 
                                  type = "partial", 
                                  variable_type = "categorical")
        var_resp$agr_profiles %>% as_tibble() %>% mutate(disease = disease)
      })
    }
  }
})

ds_pd_by_disease_dat <- ds_pd_by_disease %>% 
  janitor::clean_names() %>% 
  select(-label, -ids) 

# unique(ds_pd_by_disease_dat$x) # some cases were model_profile was averaged?

ds_pd_by_disease_dat  <- ds_pd_by_disease_dat %>% 
  mutate(x = round(as.numeric(as.character(x)))) %>% 
  distinct()

# nrow(ds_pd_by_disease_dat) == ds_pd_by_disease_dat %>% distinct(vname, x, disease) %>% nrow()

ds_pd_by_disease_dat  <- ds_pd_by_disease_dat %>% 
  mutate(x = recode(x, `1` = "disease_present", `0` = "disease_absent")) %>% 
  pivot_wider(names_from = x, values_from = yhat) %>% 
  drop_na() %>% 
  mutate(diff = disease_present - disease_absent) %>% 
  mutate(vname = recode(vname, "disease_status_lag1" = "6 months",
                        "disease_status_lag2" = "12 months",
                        "disease_status_lag3" = "18 months")) %>% 
  mutate(vname = factor(vname, levels = c("18 months", "12 months", "6 months"))) %>% 
  mutate(disease = str_replace_all(disease, "_", " ")) %>% 
  group_by(disease) %>% 
  mutate(diff_sum = sum(diff)) %>% 
  ungroup() %>% 
  mutate(disease = fct_reorder(disease, diff_sum))

ggplot(ds_pd_by_disease_dat, aes(x = disease, y = diff, fill = vname)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_manual(values = cbp1[1:3], guide = guide_legend(reverse = TRUE)) +
  coord_flip() +
  ggthemes::theme_fivethirtyeight(base_size = 22) +
  theme(legend.position = "top", axis.title = element_text(size = 14)) +
  labs(y = "Percent", x = "", fill = "") 

```
</details>

<details>
<summary>cases partial dependency of variables of interest by disease</summary>
```{r cs-lag-pd-by-disease, fig.height=10}
interesting_vars <- c("log_veterinarians_per_taxa", "log_gdp_per_capita", "cases_lag_sum_border_countries")

cs_pd_by_disease <-  map_dfr(oie_high_importance_diseases, function(disease){
  
  col <- paste("disease", disease, sep = "_")
  if(col %in% colnames(boost_dat_cases)){
    which_disease <- which(boost_dat_cases[,col] == 1)
    
    if(length(which_disease)){
      explainer_cases_sub <- DALEX::explain(
        model = boost_mod_cases_xg,
        data = boost_dat_cases[which_disease,],
        y = boost_dat_out_cases[which_disease],
        predict_function = predict_raw,
        label = "cases",
        verbose = FALSE
      )
      
      map_dfr(interesting_vars, function(var){
        var_resp <- model_profile(explainer_cases_sub, variable =  var , type = "partial", variable_type = "numerical")
        var_resp$agr_profiles %>% as_tibble() %>% mutate(disease = disease)
      })
    }
  }
})

cs_pd_by_disease_dat <- cs_pd_by_disease %>% 
  janitor::clean_names() %>% 
  select(-label, -ids, -x) %>% 
  distinct() %>% 
  group_by(vname, disease) %>% 
  summarize(diff =  (max(yhat) - min(yhat)) / min(yhat)) %>% 
  ungroup() %>% 
  mutate(vname = recode(vname, "log_veterinarians_per_taxa" = "Veterinarians per taxa",
                        "log_gdp_per_capita" = "GDP per capita",
                        "cases_lag_sum_border_countries" = "Cases in border countries\n(prev. 18 months)"
  )) %>% 
  mutate(vname = factor(vname)) %>% 
  mutate(disease = str_replace_all(disease, "_", " ")) %>% 
  group_by(disease) %>% 
  mutate(diff_sum = sum(diff)) %>% 
  ungroup() %>% 
  mutate(disease = fct_reorder(disease, diff_sum))

ggplot(cs_pd_by_disease_dat, aes(x = disease, y = diff, fill = vname)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(expand=c(0,0)) +
  scale_fill_manual(values = cbp1[c(4, 6, 7)], guide = guide_legend(reverse = TRUE)) +
  coord_flip() +
  ggthemes::theme_fivethirtyeight(base_size = 22) +
  theme(legend.position = "top", axis.title = element_text(size = 14)) +
  labs(y = "Percent (relative)", x =  "", fill = "") 
```
</details>

## Validation

```{r validation}
conn <- repeldata::repel_local_conn()
grouping_vars <- c("country_iso3c", "report_year", "report_semester", "disease", "disease_population", "taxa")

valdat <- annual_reports_animal_hosts_validation(conn) %>%
  select(all_of(grouping_vars)) %>%
  distinct() %>% 
  bootstraps(times = 1) %>% 
  pull(splits) %>% 
  magrittr::extract2(1) %>% 
  as_tibble()
```

```{r load-models}

## baseline
# forecasted_baseline <- repel_forecast(model_object = nowcast_baseline_model(),
#                                   conn = conn,
#                                   newdata = valdat)
# write_rds(forecasted_baseline, here::here("tmp/forecasted_baseline.rds"))
forecasted_baseline <- read_rds(here::here("tmp/forecasted_baseline.rds"))

scored_baseline <- repel_score(model_object = nowcast_baseline_model(),
                               augmented_data = forecasted_baseline$augmented_data,
                               predicted_cases = forecasted_baseline$predicted_cases)


## xgboost
# forecasted_boost <- repel_forecast(model_object = model_object,
#                                   conn = conn,
#                                   newdata = valdat, use_cache = TRUE)
# write_rds(forecasted_boost, here::here("tmp/forecasted_boost.rds"))
forecasted_boost <- read_rds(here::here("tmp/forecasted_boost.rds"))

scored_boost <- repel_score(model_object = model_object,
                            augmented_data = forecasted_boost$augmented_data,
                            predicted_cases = forecasted_boost$predicted_cases)

```


```{r prep}
scored_data_ds_baseline <- scored_baseline$scored_tibble %>%
  select(taxa, country_iso3c, disease, disease_status_actual, disease_status_predicted, disease_status_dirchange_actual, disease_status_dirchange_predicted) %>%
  mutate(model = "baseline")
scored_data_ds_boost <- scored_boost$scored_tibble %>%
  select(taxa, country_iso3c,  disease, disease_status_actual, disease_status_predicted, disease_status_dirchange_actual, disease_status_dirchange_predicted) %>%
  mutate(model = "xgboost")

scored_data_ds <- bind_rows(scored_data_ds_baseline, scored_data_ds_boost) %>%
  mutate_at(.vars = c("disease_status_actual", "disease_status_predicted", "disease_status_dirchange_actual", "disease_status_dirchange_predicted"), as.factor) %>%
  mutate(continent = countrycode::countrycode(country_iso3c, origin = "iso3c", destination = "continent")) %>%
  mutate(full_model = "full_model") # dummy var

metrics <- tribble(~".metric", ~"desc",
                   "accuracy", "proportion of the data that are predicted correctly",
                   "kap", "similar measure to accuracy(), but is normalized by the accuracy that would be expected by chance alone and is very useful when one or more classes have large frequency distributions.",
                   "sens", "the proportion of positive results out of the number of samples which were actually positive.",
                   "spec", "the proportion of negative results out of the number of samples which were actually negative")

ds_summary <- function(grp_var, filter_var = NULL, filter_values = NULL, truth, estimate){
  
  if(!is.null(filter_var)){
    cm_scored_mat <- scored_data_ds %>%
      filter(!!sym(filter_var) %in% filter_values)
  }else{
    cm_scored_mat <- scored_data_ds
  }
  
  cm_scored_mat <- cm_scored_mat %>%
    mutate(lab = paste(model, get(grp_var), sep = " - ")) %>%
    group_by_at(c("model", "lab", grp_var)) %>%
    conf_mat(truth = truth, estimate = estimate) %>%
    ungroup()
  
  cm_scored_summ <- cm_scored_mat %>%
    mutate(summary = map(conf_mat, summary)) %>%
    unnest(summary) %>%
    right_join(metrics)
  
  out_table <- cm_scored_summ %>%
    select(one_of(grp_var), .metric, desc, model, .estimate) %>%
    mutate(.estimate = signif(.estimate, 2)) %>%
    pivot_wider(names_from = all_of(grp_var), values_from = .estimate) %>%
    arrange(.metric)
  
  out_plots <- map(cm_scored_mat$conf_mat, autoplot, type = "heatmap")
  out_plots <- plot_grid(plotlist = out_plots, labels = cm_scored_mat$lab, label_size = 8)
  
  return(list("summary_table" = out_table, "conf_matrix" = out_plots, "raw" = cm_scored_mat))
}
```

<details>
<summary>disease status direction change confusion matrix by select diseases</summary>
```{r ds-cm-disease-dc, fig.width = 15, fig.height = 8}

ds_summary_tbl <- map_dfr(oie_high_importance_diseases, function(disease){
  
  out_disease_dc <- ds_summary("full_model", filter_var = "disease", filter_values = disease,
                               truth = "disease_status_dirchange_actual", estimate = "disease_status_dirchange_predicted")
  
  tab_ds <- out_disease_dc[[1]] %>%
    filter(.metric == "accuracy") %>%
    select(model, full_model) %>%
    pivot_wider(names_from = model, values_from = full_model)
  
  new_ds <-  out_disease_dc[[3]] %>%
    filter(model == "xgboost") %>%
    select(-model, -lab) %>%
    mutate(new_accuracy = map(conf_mat, function(cm){
      tidy(cm) %>%
        filter(name %in% c("cell_1_3", "cell_3_3")) %>%
        pivot_wider(names_from = name) %>%
        transmute(new_accuracy = cell_3_3 / (cell_3_3 + cell_1_3)) %>%
        pull(new_accuracy)
    })) %>%
    unnest(new_accuracy) %>%
    select(-conf_mat) %>%
    pull(new_accuracy)
  
  tab_ds %>%
    mutate(Disease = str_replace_all(disease,"_", " ")) %>%
    mutate(new_accuracy = new_ds)
})

ds_summary_tbl %>%
  filter(!is.nan(new_accuracy)) %>%
  arrange(-new_accuracy) %>%
  mutate_at(.vars = c("baseline", "xgboost", "new_accuracy"), .funs = ~paste0(round(100*.), "%")) %>%
  mutate(`REPEL (Overall/New Outbreaks)` = paste(xgboost, new_accuracy, sep = " / ")) %>%
  rename("Baseline Accuracy" = baseline) %>%
  select(Disease, `Baseline Accuracy`, `REPEL (Overall/New Outbreaks)`) %>% 
  kbl() %>% 
  kable_styling() 

## oie_high_importance_diseases lookup - start/end accuracy
# disease <- "foot_and_mouth_disease"
# accuracy_check <- ds_summary("full_model", filter_var = "disease", filter_values = disease,
#                                truth = "disease_status_dirchange_actual", estimate = "disease_status_dirchange_predicted")
# accuracy_check2 = accuracy_check[[3]] %>%
#   filter(model == "xgboost") %>%
#   mutate(conf_mat = map(conf_mat, tidy)) %>%
#   unnest(conf_mat) %>%
#   select(name, value) %>%
#   pivot_wider(names_from = name, values_from = value)
# 
# # outbreak starts accuracy
# accuracy_check2$cell_3_3 / (accuracy_check2$cell_1_3 + accuracy_check2$cell_2_3 + accuracy_check2$cell_3_3)
# 
# # outbreak ends accuracy
# accuracy_check2$cell_2_2 / (accuracy_check2$cell_1_2 + accuracy_check2$cell_2_2 + accuracy_check2$cell_3_2)
```
</details>

<details>
<summary>disease status direction change confusion matrix by taxa</summary>
```{r ds-cm-taxa-dc, fig.width = 15, fig.height = 8}

ds_summary_tbl <- map_dfr(unique(valdat$taxa), function(taxa){
  
  out_disease_dc <- ds_summary("full_model", filter_var = "taxa", filter_values = taxa,
                               truth = "disease_status_dirchange_actual", estimate = "disease_status_dirchange_predicted")
  
  tab_ds <- out_disease_dc[[1]] %>%
    filter(.metric == "accuracy") %>%
    select(model, full_model) %>%
    pivot_wider(names_from = model, values_from = full_model)
  
  new_ds <-  out_disease_dc[[3]] %>%
    filter(model == "xgboost") %>%
    select(-model, -lab) %>%
    mutate(new_accuracy = map(conf_mat, function(cm){
      tidy(cm) %>%
        filter(name %in% c("cell_1_3", "cell_3_3")) %>%
        pivot_wider(names_from = name) %>%
        transmute(new_accuracy = cell_3_3 / (cell_3_3 + cell_1_3)) %>%
        pull(new_accuracy)
    })) %>%
    unnest(new_accuracy) %>%
    select(-conf_mat) %>%
    pull(new_accuracy)
  
  tab_ds %>%
    mutate(Taxa = str_replace_all(taxa,"_", " ")) %>%
    mutate(new_accuracy = new_ds)
})

ds_summary_tbl %>%
  filter(!is.nan(new_accuracy)) %>%
  arrange(-new_accuracy) %>%
  mutate_at(.vars = c("baseline", "xgboost", "new_accuracy"), .funs = ~paste0(round(100*.), "%")) %>%
  mutate(`REPEL (Overall/New Outbreaks)` = paste(xgboost, new_accuracy, sep = " / ")) %>%
  rename("Baseline Accuracy" = baseline) %>%
  select(Taxa, `Baseline Accuracy`, `REPEL (Overall/New Outbreaks)`) %>% 
  kbl() %>% 
  kable_styling() 

```
</details>

<details>
<summary>disease status direction change confusion matrix by continent</summary>
```{r ds-cm-continent-dc, fig.width = 15, fig.height = 8}

ds_summary_tbl <- map_dfr(na.omit(unique(scored_data_ds$continent)), function(continent){
  
  out_disease_dc <- ds_summary("full_model", filter_var = "continent", filter_values = continent,
                               truth = "disease_status_dirchange_actual", estimate = "disease_status_dirchange_predicted")
  
  tab_ds <- out_disease_dc[[1]] %>%
    filter(.metric == "accuracy") %>%
    select(model, full_model) %>%
    pivot_wider(names_from = model, values_from = full_model)
  
  new_ds <-  out_disease_dc[[3]] %>%
    filter(model == "xgboost") %>%
    select(-model, -lab) %>%
    mutate(new_accuracy = map(conf_mat, function(cm){
      tidy(cm) %>%
        filter(name %in% c("cell_1_3", "cell_3_3")) %>%
        pivot_wider(names_from = name) %>%
        transmute(new_accuracy = cell_3_3 / (cell_3_3 + cell_1_3)) %>%
        pull(new_accuracy)
    })) %>%
    unnest(new_accuracy) %>%
    select(-conf_mat) %>%
    pull(new_accuracy)
  
  tab_ds %>%
    mutate(Continent = str_replace_all(continent,"_", " ")) %>%
    mutate(new_accuracy = new_ds)
})

 ds_summary_tbl %>%
  filter(!is.nan(new_accuracy)) %>%
  arrange(-new_accuracy) %>%
  mutate_at(.vars = c("baseline", "xgboost", "new_accuracy"), .funs = ~paste0(round(100*.), "%")) %>%
  mutate(`REPEL (Overall/New Outbreaks)` = paste(xgboost, new_accuracy, sep = " / ")) %>%
  rename("Baseline Accuracy" = baseline) %>%
  select(Continent, `Baseline Accuracy`, `REPEL (Overall/New Outbreaks)`) %>% 
  kbl() %>% 
  kable_styling() 
```
</details>

<details>
<summary>cases model stats</summary>
```{r cs-stats}
scored_data_cs_baseline <- scored_baseline$scored_tibble %>%
  select(disease, taxa, country_iso3c, cases_actual, cases_predicted, cases_dirchange_actual, cases_dirchange_predicted, cases_error) %>%
  mutate(model = "baseline") %>%
  drop_na(cases_actual)
scored_data_cs_boost <- scored_boost$scored_tibble%>%
  select(disease, taxa, country_iso3c, cases_actual, cases_predicted, cases_dirchange_actual, cases_dirchange_predicted, cases_error) %>%
  mutate(model = "xgboost") %>%
  drop_na(cases_actual)

scored_data_cs <- bind_rows(scored_data_cs_baseline, scored_data_cs_boost) %>%
  mutate(continent = countrycode::countrycode(country_iso3c, origin = "iso3c", destination = "continent")) %>%
  mutate(full_model = "full_model") # dummy var

scored_data_cs %>%
  group_by(model) %>%
  yardstick::metrics(truth = cases_actual, estimate = cases_predicted)
```
</details>

<details>
<summary>cases residuals</summary>
```{r cd-resid, fig.height=5}
scored_data_cs %>%
  ggplot(., aes(x = cases_actual, y = cases_predicted, color = log10(cases_error))) +
  geom_point(alpha = 0.5) +
  facet_wrap(model ~ .) +
  scale_y_log10(limits = c(1, max(c(scored_data_cs$cases_actual, scored_data_cs$cases_predicted)))) +
  scale_x_log10(limits = c(1, max(c(scored_data_cs$cases_actual, scored_data_cs$cases_predicted)))) +
  scale_color_viridis_c() +
  theme_bw()
```
</details>

<details>
<summary>cases residuals by disease</summary>
```{r cd-resid-disease, fig.height=10, fig.width=20}
scored_data_cs %>%
  filter(disease %in% oie_high_importance_diseases) %>% 
  ggplot(., aes(x = cases_actual, y = cases_predicted, color = log10(cases_error))) +
  geom_point(alpha = 0.5) +
  facet_grid(model ~ disease) +
  scale_y_log10(limits = c(1, max(c(scored_data_cs$cases_actual, scored_data_cs$cases_predicted)))) +
  scale_x_log10(limits = c(1, max(c(scored_data_cs$cases_actual, scored_data_cs$cases_predicted)))) +
  scale_color_viridis_c() +
  theme_bw()
```
</details>

<details>
<summary>cases residuals by taxa</summary>
```{r cd-resid-taxa, fig.height=10, fig.width=15}
scored_data_cs %>%
  ggplot(., aes(x = cases_actual, y = cases_predicted, color = log10(cases_error))) +
  geom_point(alpha = 0.5) +
  facet_grid(model ~ taxa) +
  scale_y_log10(limits = c(1, max(c(scored_data_cs$cases_actual, scored_data_cs$cases_predicted)))) +
  scale_x_log10(limits = c(1, max(c(scored_data_cs$cases_actual, scored_data_cs$cases_predicted)))) +
  scale_color_viridis_c() +
  theme_bw()
```
</details>

<details>
<summary>cases residuals by continent</summary>
```{r cd-resid-continent, fig.height=10, fig.width=15}
scored_data_cs %>%
  ggplot(., aes(x = cases_actual, y = cases_predicted, color = log10(cases_error))) +
  geom_point(alpha = 0.5) +
  facet_grid(model ~ continent) +
  scale_y_log10(limits = c(1, max(c(scored_data_cs$cases_actual, scored_data_cs$cases_predicted)))) +
  scale_x_log10(limits = c(1, max(c(scored_data_cs$cases_actual, scored_data_cs$cases_predicted)))) +
  scale_color_viridis_c() +
  theme_bw()
```
</details>

```{r disconnect}
repel_local_disconnect()
```

